\thetitle

### 2020-11-10 sun_phases/spin_qudit_tomo.tex
fixed error in Eq. 31
added comment about tightness to the main text

The assumption that observables can be estimated up to uncorrelated noise with maximal variance $\epsilon^2$, summarized by Eq.~\eqref{eq:error_assumption}, is reasonable when measurement error is dominated by experimental sources of noise.
However, this assumption breaks down when measurement error is limited by fundamental quantum (shot) noise.
We relax the assumption of Eq.~\eqref{eq:error_assumption} and derive more fundamental error bounds in Appendix \ref{sec:error}.
Therein, we find that if measurement error is dominated by shot noise, then the estimate $\tilde\rho_V$ built from $n$ independent measurements of spin projection along every axis $\v v\in V$ approximates the true state $\rho$ to within a mean squared distance $\epsilon_V^2/n$, i.e.
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2} < \f{\epsilon_V^2}{n},
  \label{eq:bound}
\end{align}
where the {\it error scale} $\epsilon_V^2$ is
\begin{align}
  \epsilon_V^2 \equiv
  \f1d \sum_{\ell>0} \p{2\ell+1} \gamma_\ell^2 \S_{V\ell}^2,
  \label{eq:scale}
\end{align}
and $\gamma_\ell$ is defined using Clebsch-Gordan coefficients:
\begin{align}
  \gamma_\ell
  \equiv \f{\max_\mu\tau_{\ell\mu} - \min_\mu\tau_{\ell\mu}}{2},
  &&
  \tau_{\ell\mu} \equiv \bk{s\mu;\ell,0|s\mu}.
\end{align}
If $d$ even or $\ell$ is odd, then $\gamma_\ell=\max_\mu\tau_{\ell\mu}$.
For comparison with the previous bound in Eq.~\eqref{eq:bound_eps}, we note that the error scale $\epsilon_V^2<\S_V^2/2$, so the previous bound still holds with the replacement $\epsilon^2\to1/2n$.

As a final comment, we note that the statistical error bound in Eq.~\eqref{eq:bound} is a ``worst case'' bound that is agnostic to the state $\rho$.
The actual statistical error for any particular state may be considerably smaller.
To write out the error in full, we identify the normalized right singular vectors $\v x^V_{\ell m}\equiv (x^V_{\ell m,1}, x^V_{\ell m,2}, \cdots)$ and corresponding singular values $M^V_{\ell m}$ of $M_{V\ell}$, and define
\begin{align}
  C_{V\ell} \equiv \sum_i C_{V\ell i} \op{i},
  &&
  C_{V\ell i} \equiv \f1{\p{M^V_{\ell m}}^2} \sum_m \abs{x_{\ell m i}}^2.
\end{align}
We can then compactly express (see Appendix \ref{sec:error})
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \f1n \sum_\ell \obk{\xi_{V\ell}|\T_{\rho\ell}},
  \label{eq:error}
\end{align}
where
\begin{align}
  \xi_{V\ell} &\equiv M_{V\ell} C_{V\ell} M_{V\ell}^\dag, \\
  \T_{\rho\ell} &\equiv \sum_{m,n}
  \cov_\rho\p{T_{\ell m}^\dag, T_{\ell n}} \op{m}{n},
\end{align}
with $\cov_\rho\p{X,Y} \equiv \bk{XY}_\rho - \bk{X}_\rho \bk{Y}_\rho$ the covariance between $X$ and $Y$.
While the error in Eq.~\eqref{eq:error} cannot be known exactly without knowing $\rho$, this error can be estimated by computing the right-hand side of Eq.~\eqref{eq:error} with the replacement $\rho\to\tilde\rho_V$.

@@

Equipped with the bound on variances in Eq.~\eqref{eq:variance_bound}, we again consider the block-diagonalized form of the measurement matrix derived in Section \ref{sec:bound}:
\begin{align}
  U M_V = \sum_\ell \op{\ell} \otimes M_{V\ell},
  &&
  M_{V\ell} = \sum_{m,\v v} D^\ell_{m,0} \op{m}{\v v}.
\end{align}
The block diagonal structure of this matrix implies that, rather than an integer index $k\in\set{1,2,\cdots,d^2}$, we can identify the singular values and right singular vectors of $UM_V$ (and equivalently $M_V$) by a block index $\ell\in\set{0,1,\cdots,d-1}$ and a sub-index $m=\set{-\ell,-\ell+1,\cdots,\ell}$.
Making the replacement $k\to\p{\ell,m}$, we can thus repeat the derivations through Eq.~\eqref{eq:bound_eps} of Section \ref{sec:bound} without the assumption of Eq.~\eqref{eq:error_assumption}, finding that
\begin{align}
  \Bbk{\abs{\tilde\rho_{\ell m}^V-\rho_{\ell m}^V}^2}
  = \sum_{i,i'} q_{\ell m i}^V \p{q_{\ell m i'}^V}^*
  \bbk{\epsilon_{i\ell} \epsilon_{i'\ell}}
  = \sum_i \abs{q_{\ell m i}^V}^2
  \bbk{\epsilon_{i\ell}^2}
  \le \f{2\ell+1}{nd} \times \gamma_\ell^2
  \times \p{1-\delta_{\ell,0}},
  \label{eq:bound_single}
\end{align}
where $i$ indexes an axis $\v v_i\in V$.
Altogether, we thus find that $n$ measurements of spin projection along every axis $\v v\in V$ allows us to construct an empirical estimate $\tilde\rho_V$ of an unknown state $\rho$ that is accurate to within a mean squared distance
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \sum_{\ell,m} \Bbk{\abs{\tilde\rho_{\ell m}^V-\rho_{\ell m}^V}^2}
  < \f1{nd} \sum_{\ell>0} \p{2\ell+1} \gamma_\ell^2 \S_{V\ell}^2.
  \label{eq:bound_SM}
\end{align}
We note that whereas the bound on $\bbk{\epsilon_{i\ell}^2}$ in Eq.~\eqref{eq:variance_bound} is tight, the bounds in Eq.~\eqref{eq:bound_single} and Eq.~\eqref{eq:bound_SM} are not: the upper bound on $\bbk{\epsilon_{i\ell}^2}$ cannot be achieved simultaneously for all axes $\v v_i$ or degrees $\ell$.
In principle, there is therefore room for improvement on the bound in Eq.~\eqref{eq:bound_SM}.

>>>>

The assumption that observables can be estimated up to uncorrelated noise with maximal variance $\epsilon^2$, summarized by Eq.~\eqref{eq:error_assumption}, is reasonable when measurement error is dominated by experimental sources of noise.
However, this assumption breaks down when measurement error is limited by fundamental quantum (shot) noise.
We relax the assumption of Eq.~\eqref{eq:error_assumption} and derive more fundamental error bounds in Appendix \ref{sec:error}.
Therein, we find that if measurement error is dominated by shot noise, then the estimate $\tilde\rho_V$ built from $n$ independent measurements of spin projection along every axis $\v v\in V$ approximates the true state $\rho$ to within a mean squared distance $\epsilon_V^2/n$, i.e.
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2} < \f{\epsilon_V^2}{n},
  \label{eq:bound}
\end{align}
where the {\it error scale} $\epsilon_V^2$ is
\begin{align}
  \epsilon_V^2 \equiv
  \f1d \sum_{\ell>0} \p{2\ell+1} \gamma_\ell^2 \S_{V\ell}^2,
  \label{eq:scale}
\end{align}
and $\gamma_\ell$ is defined using Clebsch-Gordan coefficients:
\begin{align}
  \gamma_\ell
  \equiv \f{\max_\mu\tau_{\ell\mu} - \min_\mu\tau_{\ell\mu}}{2},
  &&
  \tau_{\ell\mu} \equiv \bk{s\mu;\ell,0|s\mu}.
\end{align}
If $d$ even or $\ell$ is odd, then $\gamma_\ell=\max_\mu\tau_{\ell\mu}$.
For comparison with the previous bound in Eq.~\eqref{eq:bound_eps}, the error scale $\epsilon_V^2<\S_V^2/2$, so the previous bound still holds with the replacement $\epsilon^2\to1/2n$.
We note that the bound in Eq.~\eqref{eq:bound} is not tight, as it is acquired by bounding the statistical variance $\epsilon_{\v v\ell}^2$ in the empirical estimate of $\bk{T_{\v v\ell,0}}_\rho$ as $\bbk{\epsilon_{\v v\ell}^2}\le\frac{2\ell+1}{nd}\times\gamma_\ell^2$.
Though the individual bounds for each axis $\v v$ and degree $\ell$ are tight, these bounds cannot be achieved simultaneously.
There is therefore still room for improvement on the bound in Eq.~\eqref{eq:bound}.

As a final comment, we note that Eq.~\eqref{eq:bound} provides a pessimistic upper bound on statistical error, which can be calculated without prior knowledge of the true qudit state $\rho$.
The actual error in the reconstruction $\tilde\rho_V$ of a particular state $\rho$ may be considerably smaller.
To write out this error in full, we identify the normalized right singular vectors $\v x^V_{\ell m}\equiv (x^V_{\ell m,1}, x^V_{\ell m,2}, \cdots)$ and corresponding singular values $M^V_{\ell m}$ of $M_{V\ell}$, and define
\begin{align}
  C_{V\ell i} \equiv \sum_m \abs{\f{x_{\ell m i}}{M^V_{\ell m}}}^2,
  &&
  C_{V\ell} \equiv \sum_i C_{V\ell i} \op{\v v_i}.
\end{align}
We can then compactly express (see Appendix \ref{sec:error})
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \f1n \sum_\ell \obk{\xi_{V\ell}|\T_{\rho\ell}},
  \label{eq:error}
\end{align}
where
\begin{align}
  \xi_{V\ell} &\equiv M_{V\ell} C_{V\ell} M_{V\ell}^\dag, \\
  \T_{\rho\ell} &\equiv \sum_{m,n}
  \cov_\rho\p{T_{\ell m}^\dag, T_{\ell n}} \op{m}{n},
\end{align}
with $\cov_\rho\p{X,Y} \equiv \bk{XY}_\rho - \bk{X}_\rho \bk{Y}_\rho$.
While the error in Eq.~\eqref{eq:error} cannot be known exactly without knowing $\rho$, this error can be estimated by computing the right-hand side of Eq.~\eqref{eq:error} with the replacement $\rho\to\tilde\rho_V$.

@@

Equipped with the bound on variances in Eq.~\eqref{eq:variance_bound}, we again consider the block-diagonalized form of the measurement matrix derived in Section \ref{sec:bound}:
\begin{align}
  U M_V = \sum_\ell \op{\ell} \otimes M_{V\ell},
  &&
  M_{V\ell} = \sum_{m,\v v} D^\ell_{m,0} \op{m}{\v v}.
\end{align}
The block diagonal structure of this matrix implies that, rather than an integer index $k\in\set{1,2,\cdots,d^2}$, we can identify the singular values and right singular vectors of $UM_V$ (and equivalently $M_V$) by a block index $\ell\in\set{0,1,\cdots,d-1}$ and a sub-index $m=\set{-\ell,-\ell+1,\cdots,\ell}$.
Making the replacement $k\to\p{\ell,m}$, we can thus repeat the derivations through Eq.~\eqref{eq:bound_eps} of Section \ref{sec:bound} without the assumption of Eq.~\eqref{eq:error_assumption}, finding that
\begin{align}
  \Bbk{\abs{\tilde\rho_{\ell m}^V-\rho_{\ell m}^V}^2}
  = \sum_{i,i'} q_{\ell m i}^V \p{q_{\ell m i'}^V}^*
  \bbk{\epsilon_{i\ell} \epsilon_{i'\ell}}
  = \sum_i \abs{q_{\ell m i}^V}^2
  \bbk{\epsilon_{i\ell}^2}
  < \f{2\ell+1}{nd} \times \gamma_\ell^2
  \times \p{1-\delta_{\ell,0}},
  \label{eq:bound_single}
\end{align}
where $i$ indexes an axis $\v v_i\in V$.
Altogether, we thus find that $n$ measurements of spin projection along every axis $\v v\in V$ allows us to construct an empirical estimate $\tilde\rho_V$ of an unknown state $\rho$ that is accurate to within a mean squared distance
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \sum_{\ell,m} \Bbk{\abs{\tilde\rho_{\ell m}^V-\rho_{\ell m}^V}^2}
  < \f1{nd} \sum_{\ell>0} \p{2\ell+1} \gamma_\ell^2 \S_{V\ell}^2.
  \label{eq:bound_SM}
\end{align}
We note that whereas the bound on $\bbk{\epsilon_{i\ell}^2}$ in Eq.~\eqref{eq:variance_bound} is tight, the bounds in Eq.~\eqref{eq:bound_single} and Eq.~\eqref{eq:bound_SM} are not: the upper bound on $\bbk{\epsilon_{i\ell}^2}$ cannot be achieved simultaneously for all axes $\v v_i$ or degrees $\ell$.
In principle, there is therefore room for improvement on the bound in Eq.~\eqref{eq:bound_SM}.
